Why there are more species in several small patches vs few large patches:

### 2024-02-23 ms_SLOSS.Rmd
Updates to ms

As global biodiversity is rapidly declining due to habitat loss, it is important to determine how to protect it. The policies of many conservation agencies are still prioritizing the preservation of a single large habitat area (SL) vs. several small areas of the same size (SS), despite empirical support favouring SS. However, to date, most theories regarding this issue were based on single-species models. This work aims to explore spatial multi-species models to verify the SS > SL pattern and determine the causes of it. We used three types of models: a classical neutral, neutral with habitat selection, and a competitive hierarchical model.  All models have three parameters, namely extinction, reproduction, and dispersal distance. In the classical neutral model, species are equivalent and have the same parameters; when there is a reproduction event they send the propagule to a surrounding patch blindly, and if the patch is already occupied the propagule dies. In the neutral-with-habitat selection model, the propagules select an empty patch so their survival is ensured. In the hierarchical model, species have competitive dominance and more dominant ones can replace the less dominant with a replacement rate parameter. We use 100 species and start simulating the colonization of an empty area then we destroy a fixed amount of habitat using different regular patch sizes. The results for all models are almost always that SS retains more species than SL. The extinction of species in patches depends on the quotient between reproduction and extinction rates in relation to patch size. The only case when SL > SS is when dispersal is insufficient to connect patches, and patch size is smaller than the minimum to prevent extinction. If patch size is above the critical size to maintain local populations the SS > SL pattern is due to the sampling effect of aggregated species distribution combined with the reduction of dispersal ability of species. High dispersal rates produce homogenization of species distribution and a reduction of species richness and this is why habitat fragmentation by habitat loss retains more species when patches are smaller. The model predicts that if a set of species have more dispersal capabilities the SS effect would be lower than for species with less dispersal.

@@

Global biodiversity is rapidly declining due to, among other factors, habitat loss [@Young2016;@Chase2020;@Newbold2015]. As habitat fragmentation occurs at the same time as habitat loss, the effects of both processes are difficult to disentangle [@Haddad2015]. Landscape connectivity and organism mobility are critical features for population persistence [@Taylor1993]. Starting from a landmark paper [@Diamond1975], conservation biologists engaged in the challenging task of recommending protection measures given the environmental crisis coupled with a lack of knowledge of most of the world’s biodiversity and limited funding for conservation. In that context, early recommendations claiming to follow ecological theory derived simple thumb rules decisions to better design protected areas. However, as noted rapidly, species/area relationships do not have a clear association with fragmentation per se but mainly with overall habitat loss [@Simberloff1982].
Aiming to design better management practices, this led to a nearly 50-year Single Large vs Several Small (SLOSS) debate. While some authors argue that habitat loss is the most overwhelming effect on species loss, for others the complex factors arising from fragmentation further enhance species extinctions [@Fahrig2017;@Fletcher2018].

@@

Comparing the three versions of the model at $\lambda =2$, for higher dispersal the neutral model tends to have higher richness and Shannon diversity at higher patch sizes, followed by the hierarchical model and the neutral-with-habitat-selection model (Figure 3 & 4). At the lowest patch size, only in the neutral-with-habitat-selection have populations that can survive, because in the other model versions the individuals observed are migrants that cannot establish local populations and become extinct (See Figures S5 - S15). Thus an intermediate patch size with a maximum species richness is not observed for this model.

Examples of the final pattern of species and habitat distributions are shown in Figures S9- S32. Species spatial distributions closely match the spatial distribution previous to the habitat destruction for neutral and neutral-with-habitat-selection models. Is also clearly observed that for the hierarchical model, there is a strong dominance of the more competitive species where Shannon Diversity is very low for higher dispersal rate (Figure 4 & S23).

@@

In previous simulation studies, it was shown that fragmentation was important when the habitat amount is low, around 10% [@Hanski2015; @Rybicki2013], but it has been observed that with levels habitat around 10-20% the effect of fragmentation is positive or have an intermediate maximum [@Loke2019; @With2016], and with lower habitat amount (e.g 7%) the arrangement of patches does not matters, we obtained similar patterns with 40% habitat amount, which is near the amount used in newer simulated studies (32%) that also obtain more species richness with more fragmentation [@Rybicki2020].

The novel aspect of our model is that it considers communities that are assembled from a metacommunity in which individuals migrate to an empty habitat, and then after the number of species is stabilized the habitat is fragmented. This could be a more realistic scenario than if all species are present with initially random patterns at an already fragmented habitat [@Rybicki2020]. This could be more important the more heterogeneous was species spatial distribution pattern before fragmentation, as the cube hypothesis states.

In terms of the cube hypothesis, the between patch movement is influenced by the dispersal distance concerning patch size, as the landscape size and the percentage of habitat destruction are constant, the between-patch distance is higher for bigger patch sizes except for the biggest size that is only one patch, also the number of patches is higher for smaller patch sizes thus between patch movement will be higher. Dispersal distance also increases between patch movements, but for higher dispersal distances the sizes of the patches are less important and the relationship tends to be SS = SL.   The spreading of risk is influenced by the $\lambda$, for smaller lambda spreading the risk would be more important, but to fully incorporate the spreading the risk effect the model should include disturbances. Very low dispersal distance favours still favours combined with low $\lambda$ favours SL > SS up to a threshold after which the patch size is not important. The species' spatial clumping is influenced by the dispersal distance, higher dispersal distance produces less clumpines but this is cofounded with patch movement, so the effect is difficult to disentangle. Thus the cube hypothesis is generally compatible with the results of our model.

>>>>

As global biodiversity is rapidly declining due to habitat loss, it is important to determine how to protect it. The policies of many conservation agencies are still prioritizing the preservation of a single large habitat area (SL) vs. several small areas of the same size (SS), despite empirical support favouring SS. However, to date, most theories regarding this issue were based on single-species models. This work aims to explore spatial multi-species models to verify the SS > SL pattern and determine its causes. We used three types of models: a classical neutral, neutral with habitat selection, and a competitive hierarchical model.  All models have three parameters, namely extinction, reproduction, and dispersal distance. In the classical neutral model, species are equivalent and have the same parameters; when there is a reproduction event they send the propagule to a surrounding patch blindly, and if the patch is already occupied the propagule dies. In the neutral-with-habitat selection model, the propagules select an empty patch so their survival is ensured. In the hierarchical model, species have competitive dominance and more dominant ones can replace the less dominant with a replacement rate parameter. We use 100 species and start simulating the colonization of an empty area then we destroy a fixed amount of habitat using different regular patch sizes. The results for all models are almost always that SS retains more species than SL. The extinction of species in patches depends on the quotient between reproduction and extinction rates in relation to patch size. The only case when SL > SS is when dispersal is insufficient to connect patches, and patch size is smaller than the minimum to prevent extinction. If patch size is above the critical size to maintain local populations the SS > SL pattern is due to the sampling effect of aggregated species distribution combined with the reduction of dispersal ability of species. High dispersal rates produce homogenization of species distribution and a reduction of species richness and this is why habitat fragmentation by habitat loss retains more species when patches are smaller. The model predicts that if a set of species have more dispersal capabilities the SS effect would be lower than for species with less dispersal.

@@

Global biodiversity is rapidly declining due to, among other factors, habitat loss [@Young2016;@Chase2020;@Newbold2015]. As habitat fragmentation occurs at the same time as habitat loss, the effects of both processes are difficult to disentangle [@Haddad2015]. Landscape connectivity and organism mobility are critical features for population persistence [@Taylor1993]. Starting from a landmark paper [@Diamond1975], conservation biologists engaged in the challenging task of recommending protection measures given the environmental crisis coupled with a lack of knowledge of most of the world’s biodiversity and limited funding for conservation. In that context, early recommendations claiming to follow ecological theory derived simple thumb rules to better design protected areas. However, as noted rapidly, species/area relationships do not have a clear association with fragmentation per se but mainly with overall habitat loss [@Simberloff1982].
Aiming to design better management practices, this led to a nearly 50-year Single Large vs Several Small (SLOSS) debate. While some authors argue that habitat loss is the most overwhelming effect on species loss, for others the complex factors arising from fragmentation further enhance species extinctions [@Fahrig2017;@Fletcher2018].

@@

Comparing the three versions of the model at $\lambda =2$, for higher dispersal the neutral model tends to have higher richness and Shannon diversity at higher patch sizes, followed by the hierarchical model and the neutral-with-habitat-selection model (Figures 3 & 4). At the lowest patch size, only in the neutral-with-habitat-selection have populations that can survive, because in the other model versions the individuals observed are migrants that cannot establish local populations and become extinct (See Figures S5 - S15). Thus an intermediate patch size with a maximum species richness is not observed for this model.

Examples of the final pattern of species and habitat distributions are shown in Figures S9- S32. Species spatial distributions closely match the spatial distribution previous to the habitat destruction for neutral and neutral-with-habitat-selection models. Is also clearly observed that for the hierarchical model, there is a strong dominance of the more competitive species where Shannon Diversity is very low for a higher dispersal rate (Figure 4 & S23).

@@

In previous simulation studies, it was shown that fragmentation was important when the habitat amount is low, around 10% [@Hanski2015; @Rybicki2013], but it has been observed that with habitat levels around 10-20% the effect of fragmentation is positive or have an intermediate maximum [@Loke2019; @With2016], and with lower habitat amount (e.g 7%) the arrangement of patches does not matters, we obtained similar patterns with 40% habitat amount, which is near the amount used in newer simulated studies (32%) that also obtain more species richness with more fragmentation [@Rybicki2020].

The novel aspect of our model is that it considers communities that are assembled from a metacommunity in which individuals migrate to an empty habitat. After the number of species is stabilized the habitat is fragmented. This scenario is deemed more realistic than having all species initially present with random patterns in an already fragmented habitat  [@Rybicki2020]. This could be more important the more heterogeneous was species spatial distribution pattern before fragmentation, as the cube hypothesis states.

In terms of the cube hypothesis, low between-patch movement, minimal impact of spreading-of-risk on population dynamics, and low species spatial clumping are needed for SL > SS. In our model dispersal distance increases between patch movements, but for higher dispersal distances the sizes of the patches are less important and the relationship tends to be SS = SL. Spreading-of-risk is influenced by $\lambda$, which represents the relationship between the growth and death rates. For smaller $\lambda$, spreading-of-risk becomes more significant because the probability of local extinction is higher. A very low dispersal distance combined with low $\lambda$ favors SL > SS up to a threshold in patch size, beyond which SL = SS. The spatial clumping of species is influenced by the dispersal distance; a higher dispersal distance produces less clumpiness. However, this is confounded with patch movement, making it challenging to disentangle the effect. Therefore, the results of our model are generally compatible with the cube hypothesis.
